---
title: An R Markdown document converted from "R_Intro.ipynb"
output: html_document
---


# R Introduction

-----------------------

BIO401-01/598-02 
 
Mar. 17 2021




 R is a script language for statistical data analysis and manipulation. 
 
 home page : https://cran.r-project.org/

## How to run 

1. batch mode

    $ R CMD BATCH script.R
    
help for the batch mode 

    $ R CMD command --help 

for example 

    $ R CMD INSTALL --help 

2. interactive mode

console : ">" as the R prompt

common IDE (integrated development environment) for R 

- R studio 
- Jupyter-notebook 
- ESS(emacs speaks statistics) 
- R commander, etc.

## Installing packages

    > install.packages("package_name")
        
    > install.packages(c("ggplot2","sp","raster"))

## Interfacing with the system

    > dir() #  show files in the current directory

    > getwd() #  is asking for the current working directory

## Getting help

R provides help with function and commands. On-line help gives useful 
information as well. Getting used to R help is a key to successful 
statistical modelling. The online help can be accessed in HTML format by 
typing:

    > help.start()
    
A keyword search is possible using the Search Engine and Keywords link.
You can also use the help() or ? functions.

    > help(dir)

or 

    > ? dir    

## As a caculator

```{r}
3 + 5 
```

```{r}
sin(pi/2)
```

```{r}
2^3
```

```{r}
5/3
```

```{r}
5%%3
```

```{r}
5%/%3
```

## Logic Operations

- AND operator &
- OR operator |
- NOT operator !

```{r}
12 > 5 & 12 < 15
```

```{r}
! TRUE
```

```{r}
TRUE | FALSE
```

## Data structures 

R objects
  
The entities R operates on are technically known as  objects. 
Examples are "vectors of numeric (real)" or "complex values", "vectors of 
logical" values and "vectors of character strings". 
These are known as  “atomic”  structures since their components are all of 
the same type, or mode, namely numeric, complex, logical, character and raw.
R also operates on objects called "lists", which are of mode list. 
These are ordered sequences of objects which individually can be of any mode. 
Lists are known as  “recursive”  rather than atomic structures since their 
components can themselves be lists in their own right.
The other recursive structures are those of mode function and  expression. 

By the "mode" of an object we mean the basic type of its fundamental 
constituents. This is a special case of a  “property”  of an object. Another
property of every object is its "length." The functions mode(object) and 
length(object) can be used 
to find out the mode and length of any defined structure 10.

Further properties of an object are usually provided by attributes(object). 
Because of this, mode and length are also called “intrinsic attributes” of an object. 
For example, if z is a complex vector of length 100, then in an expression mode(z) 
is the character string "complex" and length(z) is 100.

### vectors : the heart of R

Vectors are combinations of scalars in a string structure. Vectors must have 
all values of the same mode. Thus any given vector must be unambiguously 
either logical, numeric, complex, character or raw. (The only apparent 
exception to this rule is the special “value” listed as NA for quantities not
available, but in fact there are several types of NA). Note that a vector can
be empty and still have a mode. For example the empty character string vector
is listed as character(0) and the empty numeric vector as numeric(0). 

    > help(vector)

```{r}
# "<-" is the  assignment operator in R.
# "c" stands for concatenate. 

x <- c(1,5,8)
```

```{r}
# subsetting operation
# "[]" used to retrieve vector elements 
# ":" used to give a range for retrieval
x[1]
x[2:3]
```

```{r}
length(x)
```

```{r}
mode(x)
```

```{r}
# scalar not really exist in R 
# individual numbers actually single element vectors 

x <- 3.5
x
x[1]
```

Two common functions we can use to generate a sequence.  

The seq() function "seq(from = number, to = number, by = number)" allow to 
create a vector starting from a value to another by a defined increment:
    
The replicate function  "rep(x,times)"  enables you to replicate a vector 
several times in a more complex vector. Calculations can be included to 
form vectors as well and functions can be combined in the same command.

```{r}
1:10
```

```{r}
seq(1,3,0.25)
```

```{r}
one2three <- 1:3 
rep(one2three,5) 
```

### Missing Values

In some cases the components of a vector or of an R object more in general, 
may not be completely known. When an element or value is “not available” 
or a “missing value” in the statistical sense, a place within a vector may
be reserved for it by assigning it the special value NA. Any operation on 
an NA becomes an NA. 

The function is.na(x)  gives a logical vector of the same size as x with 
value TRUE if and only if the corresponding element in x is NA.

Functions : dealing with NAs

na.fail returns the object if it does not contain any missing values, and signals an error otherwise. 

na.omit returns the object with incomplete cases removed. 

na.pass returns the object unchanged.

```{r}
z <- c(1:3,NA)
ind <- is.na(z)
ind
```

There is a second kind of “missing” values which are produced by numerical 
computation, the so-called Not a Number,  NaN , values. 

Examples are 0/0 
or Inf - Inf which both give NaN since the result cannot be defined sensibly.

```{r}
Inf-Inf
0/0
```

is.na(xx) is TRUE both for NA and NaN values. To differentiate 
these, is.nan(xx) is only TRUE for NaNs. Missing values are sometimes printed
as <NA> when character vectors are printed without quotes. 

```{r}
z <- c(1:3,NA,0/0)
is.not.available <- is.na(z) 
is.not.a.number <-is.nan(z)

is.not.a.number
is.not.available 
```

On the other hand, NULL represent the value does not exit. 

```{r}
x <- c(1,2,3,NA)
mean(x)
```

```{r}
mean(x,na.rm=TRUE)
```

```{r}
y <- c(1,2,3,NULL)
mean(y)
```

```{r}
na.fail(x)
```

```{r}
na.fail(y)
```

#### Exercise

creat a sequence as below and compute its mean

```
10 20 30 10 20 30 50 50 50 NA NA NA 10 10 10  
```

### Working with strings

character strings : single element vectors 

mode : check datatype 

```{r}
x <- "hello"
mode(x)
length(x)
```

```{r}
# concatenate strings 

paste("hello","world")
```

```{r}
# if you don't want any space between the strings, you can use paste0

paste0 ("hello","world")
```

```{r}
# substr is used to get parts of a string 

substr("Hello world", 6,11)
```

```{r}
# grep : retrieval based on certain pattern 

timespan <- c("day","month","year")
grep('y', timespan)
```

```{r}
# ^ : start with

grep('^m', timespan)
```

```{r}
numeric.vector <- c(rep(c (5*10:1, 5, 6), 2))
numeric.vector
```

```{r}
character.vector  <- as.character(numeric.vector)
character.vector
```

#### Exercise

retreive the position of the 2nd "to" in the quatation in the sentence below, replace it with "T" and reprint the whole sentence.

```
“All we have to decide is what to do with the time that is given us.”

― J.R.R. Tolkien, The Fellowship of the Ring
```

### Factors

A factor is a variable to represent categories of a set. They can created using the function as.factor(). 

```{r}
countries <- c("Korea", "China", "UK", "UK", "USA", "Japan", "Korea")
as.factor(countries)
```

```{r}
# factor can work on numbers also. 
fnum <- c(1:3,2:5,1:8)
as.factor(fnum)
```

### Set Operations 

    union(x,y) 
    intersect(x,y)
    setdiff(x,y) : all elements of x NOT in y
    setequal(x,y)
    c %in% x : membership testing 

```{r}
x <- c(1, 3, 5, 8, 10)
y <- c(3, 4, 5, 6, 7)
```

```{r}
union(x,y)
intersect(x,y)
setdiff(x,y)
setequal(x,y)
```

```{r}
7 %in% x
```

### List : contains that can hold different types 

Lists are a general form of vector in which the various elements need not be 
of the same type, and are often themselves vectors or lists. Lists provide a 
convenient way to return the results of a statistical computation.

```{r}
x <- list(w=2, v="GIS")
x
```

```{r}
# $ sign used to access list elements by names
x$w
```

```{r}
mode(x)
mode(c(x$w,x$v))
```

```{r}
x <- c(2,"GIS")
mode(x)
mode(x[1])
```

```{r}
# show internal datasets 

data()
```

```{r}
hn <- hist(Nile)
```

```{r}
print (hn)
```

```{r}
# str : structure
str(hn)
```

```{r}
mode(hn)
```

```{r}
# summary is a generic function in R. It outputs a concise and more friendly representation of a list. 
summary(hn)
```

Matrices

Matrices, or more generally arrays, are multi-dimensional generalizations of 
vectors. In fact, they are vectors that can be indexed by two or more indices
and will be printed in a special way.

The matrix() function creates a matrix from the given set of values. We use 
the matrix(x, nrow=, ncol=) function to set the matrix cell values, the 
number of rows and the number of columns. We can use the colnames() and 
rownames() functions to set the column and row names of the matrix-like 
object.

```{r}
matrix(data = NA, nrow = 2, ncol = 3) 
```

```{r}
example.matrix <- matrix(1:6,2,3)
example.matrix
```

```{r}
# data distributed by columns by default, can be changed using the keyword byrow 
example.matrix <- matrix(1:6,2,3, byrow=TRUE)
example.matrix
```

```{r}
# retrieval by rows
example.matrix[1,]
```

```{r}
# retrival by columns
example.matrix[,2]
```

```{r}
# changing values

example.matrix[1,] <- 1:3
example.matrix[2,] <- c(5,10,4)
example.matrix
```

```{r}
# apply () function is a machanism to apply a function across a vector. 

# by column
apply(example.matrix,2,sum)
```

```{r}
# by row
apply(example.matrix,1,sum)
```

```{r}
matrix.head <- c("col a","col b","column c")
matrix.side <- c("first raw","second raw")
```

```{r}
colnames(example.matrix) = matrix.head
rownames(example.matrix) = matrix.side
example.matrix
```

```{r}
# The structure function str(object.name) informs you of the structure of a
# specific object

str(example.matrix)
```

```{r}
# combine matrix using rbind (by row) and cbind (by column)

matrix1 <- matrix(1:6,2,3)
matrix2 <- matrix(11:16,2,3)
rbind(matrix1,matrix2)
cbind(matrix1,matrix2)
```

### Exercise 

create a matrix with the following elements and then change the first column values to all 0.

```
1 1 1 
2 2 2 
3 3 3 
```

Arrays

An array can be considered a multiple subscripted collection of data 
entries, for example numeric. R allows simple facilities for creating 
and handling arrays, and in particular the special case of matrices. 

As well as giving a vector structure a dim attribute, arrays can be 
constructed from vectors by the array function, which has the form 
array(data_vector, dim_vector)

```{r}
Z <- array(1:24, c(3,4,2))
Z
```

```{r}
str(Z)
```

## list and delete R objects 

The list function ls() outputs a list of existing R objects.

rm() removes the object.

```{r}
ls()
```

```{r}
rm(Z)
ls()
```

## Data Frames

Data frames are matrix-like structures, in which the columns can be 
of different types. Think of data frames as  data matrices  with one row per 
observational unit but with (possibly) both numerical and categorical 
variables. Many experiments are best described by data frames: the treatments
are categorical but the response is numeric. 

As a result R dataframes are tightly coupled collections of variables which 
share many of the properties of matrices and of lists. Data frames are used 
as the fundamental data structure by most of R's modeling software.

A data frame is a list with class "data.frame". There are restrictions on 
lists that may be made into data frames, namely :

The components must be vectors (numeric, character, or logical), factors, 
numeric matrices, lists, or other data frames.
Matrices, lists, and data frames provide as many variables to the new data 
frame as they have columns, elements, or variables, respectively.
Numeric vectors, logicals and factors are included, and character vectors 
are coerced to be factors, whose levels are the unique values appearing in
the vector.
Vector structures appearing as variables of the data frame must all have the
same length, and matrix structures must all have the same row size. 

### Dataframe construction 

```{r}
my.data.frame = data.frame(v = 1:4, ch = c("a", "b", "c", "d"), n = 10)
my.data.frame
```

```{r}
# Or:

my.data.frame = data.frame(vector = 1:4,
                           character = c("a", "b", "c", "d"),
                           const.vector = 10,
                           row.names =c("data1", "data2", "data3", "data4"))
my.data.frame
```

### Data selection and manipulation

```{r}
# You can extract data from dataframes using the    [ [    ] ]  and  $  sign:

my.data.frame[["character"]]

my.data.frame[[2]]
```

```{r}
# Call the 3rd value of the character vector:

my.data.frame[[2]][3]
```

```{r}
# Or using the $ syntax:
  
my.data.frame$vector
  
my.data.frame$character[2:3]
```

```{r}
# You can add single arguments to a data frame, query information, select and 
# manipulate arguments or single values from a dataframe 
  
my.data.frame$new

my.data.frame$new = c(10,11,20,40)
my.data.frame
```

```{r}
# length(object.name) returns the number of elements in an object such as 
# matrix vector or dataframes:
  
length(my.data.frame$new) 
```

```{r}
# which(object.name) and which.max(object.name) return the index of a specific
# or of the greatest element of an object
    
which.max(my.data.frame$new) 

which(my.data.frame$new == 20) 
```

```{r}
# max(object.name) returns the value of the greatest element
  
max(my.data.frame$new) 
```

```{r}
# sort(object.name) sort from small to big 
  
sort(my.data.frame$new) 
```

```{r}
# rev(object.name) sorts from big to small
  
rev(sort(my.data.frame$new)) 
```

```{r}
# subset(object.name, ...) returns a selection of an R-object with respect to 
# criteria (typically comparisons: x$V1 < 10). 
  
subset(my.data.frame, my.data.frame$new == 20)
```

```{r}
# If the R-object is a data frame,
# the option select gives the variables to be kept or dropped using a minus 
# sign

my.data.frame[-1]
```

## Reading and Writing Files

read.table("filename")** Reads a file in table format and creates a data 
frame from it, with cases corresponding to lines and variables to fields 
within the file. The default separator  sep=""  is any whitespace. 
You might need  sep=","  or  ";"  and so on. 
Use  header=TRUE  to read the first line as a header of column names. 
The **as.is=TRUE** specification is used to prevent character vectors from
being converted to factors. 

The comment.char="" specification is used to prevent "#" from being 
interpreted as a comment and use "skip=n" to skip n lines before reading 
data. For more details:
?read.table

read.csv("filename")  is set to read comma separated files. Example usage is:
read.csv(file.name, header = TRUE, sep = ",", quote="\"", dec=".", 
fill =TRUE,  comment.char="", ...)

read.delim("filename") is used for reading tab-delimited files

read.fwf() reads a table of fixed width formatted data into a ’data.frame’.
Widths is an integer vector, giving the widths of the fixed-width fields

write.table(Robj, "filename", sep=",", row.names=FALSE, quote=FALSE)

```{r}
df <- read.csv("./geodata/shp/point_stat.csv")
```

```{r}
head(df)
```

```{r}
# subsetting 
dfSel <- df[which(df$mean>7),]
dfSel
```

```{r}
write.csv(dfSel,"./geodata/shp/point_stat_sel.csv", row.names=FALSE)
```

```{r}
df <- read.csv("./geodata/shp/point_stat_sel.csv")
df
```

### Exercise 

Print the data rows with the column min value less than 1. Do NOT use the "which" function in R. 

## Functions

Functions are themselves objects in R which can be stored in the project's 
workspace. This provides a simple and convenient way to extend R.  

Usage: in writing your own function you provide one or more arguments or 
names for the function, an expression and a value 
is produced equal to the output function result.

function(arglist) expr   function definition  
return(value) 

```{r}
# Example

myfunction <- function(x) x^5
myfunction(3)
```

```{r}
# oddcount function 

oddcount <- function(x){
    k <- 0 
    for (i in x) {
        if (i %% 2 == 1) k <- k + 1
    }
    return (k)
}
```

```{r}
oddcount(c(1,3,5,7))
```

```{r}
oddcount(c(1,3,5,8))
```

### break and next 
    break : break out of a loop 
    next : skip a step 

```{r}
for (i in 1:10){
    if (i %in% c(1,3,5,7)) {
        next
    }
    else if (i > 8){
        break
    }        
    print (i)
}
```

### Exercise 

write a function to test if a number is a prime number, a natural number that is NOT a multiplication of two smaller numbers. 

## Basic Graphs

The plot() function forms the foundation for much of R based graphing operations. 

```{r}
plot(c(1,3,5,7),c(2,4,6,8))
```

```{r}
x <- seq(0,10,length.out=100)
y <- exp(-x)
```

```{r}
plot(x,y)
```

```{r}
plot(x,y,type="l",lwd=3)
```

```{r}
plot(x,y,type="l",lwd=3, xlab="time", cex.lab=1.5,cex.axis=1.5)
grid(lwd=3)
```

## Assignment 

- individual effort or collaboration by maximum two people 
- if in collaboration, please specify the contribution of each contributor 
- please turn in a set of .pdf and .ipynb files
- please label the question number before giving the answer

Q1. Read in file "./geodata/shp/point_stat.csv" and output the rows that their values in the "stdev" column are less than those in their "min" column and their "mean" values are all greater than 7. (20%)


Q2. Create a 3x3 matrix with element values in the range of 1 to 9 and then change the diagonal elements to 0. (20%)

```
1	4	7           0	4	7
2	5	8    ==>    2	0	8
3	6	9           3	6	0
```


Q3. Create a dataframe to store the calendar of March 2021 with the day of the week as the column name. (20%) 

```
Mo Tu We Th Fr Sa Su 
 1  2  3  4  5  6  7 
 8  9 10 11 12 13 14 
15 16 17 18 19 20 21 
22 23 24 25 26 27 28 
29 30 31
```
QB (bonus) (10%)

According to Gregorian calendar, leap years can be determined by the criteria below. 
```
Years that are divisible by 4 and not divisible by 100 are leap years. 
Years that are divisible by 100 but not divisible by 400 are NOT leap years. 
Years that are divisible by 400 are leap years. 
```
Please write a R script to print all the leap years between 1800 and 2020 into a file named "leap_years.dat" and calculate the ratio of the number of leap years over this period of time.

